\textbf{Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces}

### 2025-06-13 paper/manuscript.qmd
elements of discussion

The same analysis was run for Beauty, which indicated the following differences: 
**ANA: can you report the main differences?**

@@

The findings of Study 2 replicate and extend those of Study 1, supporting the idea that facial attractiveness significantly shapes participants' beliefs about reality beliefs pertaining to human faces. This effect was particularly strong and robust among male participants. For females, the effect was weaker in Study 1 but more pronounced in Study 2, which might partly reflect increased statistical power.

Another possibility, however, is that the second sample's experimental manipulation introduced conflicting cues (e.g., faces labelled "AI-generated") that could have disrupted intuitive heuristics, thereby increasing reliance on attractiveness as a diagnostic cue, particularly in ambiguous trials.
<!-- This could be tested by analyzing believed vs. non-believed items -->

The fact that the effect was less obvious in women in the sample 1 might be explained with power (less relevant items were included for women).

@@

We demonstrated the flexibility of the CHOCO model, being usable in analog scales as well as discrete ratings with a mid-point.

Likert-scales (with a limited number of discrete options) are commonly used in psychology, and often modeled as continuous.
While this fact has spurred its own polarized debate (REFS), our position is that the model used should ideally reflect the data *generating* mechanism rather than necessarily focus on the *observed* data^[Ideally, the data recording method should be aligned with the assumed generating mechanism, which is an experiment design issue rather than a statistical one.] - although the former cannot always be easily inferred or known. 
For the data of Sample 2, we held the assumption that the underlying latent distribution was CHOCO distributed (as there is no reason to assume it would be different from Sample 1), treated as continuous CHOCO-distributed data, despite the fact that it was collected on a 7-point Likert scale. 
The motivation was to see if the model could extrapolate and reliably estimate all the parameters based on a limited number of data points but means in our case that, effectively, each Beta distribution was extrapolated from the frequency of only two unique values (or 3 for the no-extreme version). 
While this is far from ideal, the CHOCO model still provided a better fit to the data than alternative approaches. 
**TODO** However, it might have lead to a lesser ability to detect finer grain changes, including U-shaped relationships. Having 7-points (or 6 with no mid-value) is the strict minimum to use the CHOCO model.

@@

Future studies should assess psychometric property liek the minimum amount of data required to estimate and recover all parameters reliably.

>>>>

The same analysis was run for Beauty, which indicated the following differences: 
**ANA: can you check and write down the main differences?**

@@

This study provides further evidence that attractive faces are more likely to be judged as real (*vs.* AI-generated), This effect was particularly strong and robust among male participants. For females, the effect was weaker in sample 1 but more pronounced in sample 2, which might partly reflect increased statistical power. An alternative possibility is that the second sample's experimental manipulation introduced conflicting cues (e.g., faces labelled "AI-generated") that could have disrupted intuitive heuristics, thereby increasing reliance on attractiveness as a diagnostic cue, particularly in ambiguous trials.
<!-- This could be tested by analyzing believed vs. non-believed items -->

<!-- Paragraph about potential mechanisms -->
However, the mechanisms that underlie why attractiveness predicts reality judgments remains unclear.
One possibility is that this positive link reflects a broader bias toward attributing positive qualities to attractive individuals, in-line with the substantial body of literature on the "beauty-is-good" stereotype, whereby attractive faces are judged as more trustworthy [@eagly1991beautiful], warmer and more approachable [@fiske2002model], and perceived as more authentic and sincere [@little2021facial]. These traits may implicitly overlap with qualities ascribed to "real" humans. In this sense, reality judgments could be an extension of existing social heuristics that conflate visual appeal with genuineness.

Alternatively, the effect might be primarily driven by attention. Attractive faces tend to capture and hold visual attention more effectively [@nakamura2014attractive], which could lead to a deeper processing. 
This greater accumulation of evidence could influence the bias systematically towards a particular category ("reality") or perhaps, towards the "true" category of the stimulus (which we cannot delineate from our data as all our stimuli were actually real photographs). Alternatively, such attentional bias may boost perceptual fluency and subjective certainty, that would be predominantly reflected in the confidence parameter of the CHOCO model. The attention hypothesis could be formally tested via incorporating a reality judgment task of real and actual AI-generated faces within attentional paradigms to see whether the confidence and bias changes as a function of attentional engagement. 
<!-- This could be tested via an attention (ANT) paradigm. -->

Importantly for the scope of this study, the CHOCO model allowed us to dissect this effect in greater detail compared to traditional analyses, revealing that attractiveness not only shifted the choice toward "real" judgments but also modulated the confidence and extremity of those judgments. With the second sample, we demonstrated the flexibility of this model by fitting it to discrete ratings with a mid-point - providing potential insight into epistemic uncertainty in decision-making tasks.

However, one might be concerned about the application of the CHOCO model (a mixture of two continuous Beta distributions) to ordinal data in Sample 2 (where a 7-point Likert scale was used). Ordinal ratings (limited number of discrete options represented numerically by integers) are commonly used in psychology, and often modeled as continuous. While this fact has spurred its own polarized debate (REFS), our position is that the model used should ideally reflect the data *generating* mechanism rather than necessarily focus on the *observed* data^[Ideally, the data recording method should be aligned with the assumed generating mechanism, which is an experiment design issue rather than a statistical one.] - although the former cannot always be easily inferred or known. 
For the data of Sample 2, we held the assumption that the underlying latent distribution was CHOCO distributed (as there is no reason to assume it would be different from Sample 1), treated as continuous CHOCO-distributed data, despite the fact that it was collected on a 7-point scale. 
The motivation was to see if the model could extrapolate and reliably estimate all the parameters based on a limited number of data points but means in our case that, effectively, each Beta distribution was extrapolated from the frequency of only three unique values. 
While this is far from ideal, the CHOCO model still proved insightful, although it might have impacted its ability detect finer-grain changes, including quadratic relationships.

<!-- Interestingly, in Sample 2, prior labels (“AI-generated” vs. “Photograph”) had a modest but significant influence on beliefs—particularly for females—suggesting that top-down expectations can modulate reality perception. This effect was small but detectable, and illustrates how pre-existing contextual information can shape subjective reality beliefs, possibly interacting with heuristic features such as attractiveness. -->

@@

@@

Future studies should assess psychometric property like the minimum amount of data required to estimate and recover all parameters reliably. As well as the minimum amount of unique value (for Likert scales) requires to detect weak or non-linearly shaped effects.

### 2025-06-13 paper/manuscript.tex
elements of discussion

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, colback=white, bottomrule=.15mm, breakable, toprule=.15mm, rightrule=.15mm, arc=.35mm, left=2mm, leftrule=.75mm]

@@

The findings of Study 2 replicate and extend those of Study 1,
supporting the idea that facial attractiveness significantly shapes
participants' beliefs about reality beliefs pertaining to human faces.
This effect was particularly strong and robust among male participants.
For females, the effect was weaker in Study 1 but more pronounced in
Study 2, which might partly reflect increased statistical power.

Another possibility, however, is that the second sample's experimental
manipulation introduced conflicting cues (e.g., faces labelled
``AI-generated'') that could have disrupted intuitive heuristics,
thereby increasing reliance on attractiveness as a diagnostic cue,
particularly in ambiguous trials.

The fact that the effect was less obvious in women in the sample 1 might
be explained with power (less relevant items were included for women).

@@

We demonstrated the flexibility of the CHOCO model, being usable in
analog scales as well as discrete ratings with a mid-point.

Likert-scales (with a limited number of discrete options) are commonly
used in psychology, and often modeled as continuous. While this fact has
spurred its own polarized debate (REFS), our position is that the model
used should ideally reflect the data \emph{generating} mechanism rather
than necessarily focus on the \emph{observed} data\footnote{Ideally, the
  data recording method should be aligned with the assumed generating
  mechanism, which is an experiment design issue rather than a
  statistical one.} - although the former cannot always be easily
inferred or known. For the data of Sample 2, we held the assumption that
the underlying latent distribution was CHOCO distributed (as there is no
reason to assume it would be different from Sample 1), treated as
continuous CHOCO-distributed data, despite the fact that it was
collected on a 7-point Likert scale. The motivation was to see if the
model could extrapolate and reliably estimate all the parameters based
on a limited number of data points but means in our case that,
effectively, each Beta distribution was extrapolated from the frequency
of only two unique values (or 3 for the no-extreme version). While this
is far from ideal, the CHOCO model still provided a better fit to the
data than alternative approaches. \textbf{TODO} However, it might have
lead to a lesser ability to detect finer grain changes, including
U-shaped relationships. Having 7-points (or 6 with no mid-value) is the
strict minimum to use the CHOCO model.

@@

@@

>>>>

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, breakable, colback=white, opacityback=0, rightrule=.15mm, arc=.35mm, bottomrule=.15mm, left=2mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, leftrule=.75mm]

@@

This study provides further evidence that attractive faces are more
likely to be judged as real (vs.~AI-generated), This effect was
particularly strong and robust among male participants. For females, the
effect was weaker in sample 1 but more pronounced in sample 2, which
might partly reflect increased statistical power. An alternative
possibility is that the second sample's experimental manipulation
introduced conflicting cues (e.g., faces labelled ``AI-generated'') that
could have disrupted intuitive heuristics, thereby increasing reliance
on attractiveness as a diagnostic cue, particularly in ambiguous trials.

However, the mechanisms that underlie why attractiveness predicts
reality judgments remains unclear. One possibility is that this positive
link reflects a broader bias toward attributing positive qualities to
attractive individuals, in-line with the substantial body of literature
on the ``beauty-is-good'' stereotype, whereby attractive faces are
judged as more trustworthy (\citeproc{ref-eagly1991beautiful}{Eagly et
al., 1991}), warmer and more approachable
(\citeproc{ref-fiske2002model}{Fiske et al., 2002}), and perceived as
more authentic and sincere (\citeproc{ref-little2021facial}{Little,
2021}). These traits may implicitly overlap with qualities ascribed to
``real'' humans. In this sense, reality judgments could be an extension
of existing social heuristics that conflate visual appeal with
genuineness.

Alternatively, the effect might be primarily driven by attention.
Attractive faces tend to capture and hold visual attention more
effectively (\citeproc{ref-nakamura2014attractive}{Nakamura \& Kawabata,
2014}), which could lead to a deeper processing. This greater
accumulation of evidence could influence the bias systematically towards
a particular category (``reality'') or perhaps, towards the ``true''
category of the stimulus (which we cannot delineate from our data as all
our stimuli were actually real photographs). Alternatively, such
attentional bias may boost perceptual fluency and subjective certainty,
that would be predominantly reflected in the confidence parameter of the
CHOCO model. The attention hypothesis could be formally tested via
incorporating a reality judgment task of real and actual AI-generated
faces within attentional paradigms to see whether the confidence and
bias changes as a function of attentional engagement.

Importantly for the scope of this study, the CHOCO model allowed us to
dissect this effect in greater detail compared to traditional analyses,
revealing that attractiveness not only shifted the choice toward
``real'' judgments but also modulated the confidence and extremity of
those judgments. With the second sample, we demonstrated the flexibility
of this model by fitting it to discrete ratings with a mid-point -
providing potential insight into epistemic uncertainty in
decision-making tasks.

However, one might be concerned about the application of a continuous
distribution (the CHOCO mixture of two Beta distributions) to ordinal
data (in Sample 2, where a 7-point Likert scale was used), as it
translated effectively to only three unique values used to estimate a
Beta distribution. Ordinal ratings (limited number of discrete options
represented numerically by integers) are commonly used in psychology,
and often modeled as continuous. While this fact has spurred its own
polarized debate (REFS), our position is that the model used should
ideally reflect the data \emph{generating} mechanism rather than
necessarily focus on the \emph{observed} data\footnote{Ideally, the data
  recording method should be aligned with the assumed generating
  mechanism, which is an experiment design issue rather than a
  statistical one.} - although the former cannot always be easily
inferred or known. For the data of Sample 2, we held the assumption that
the underlying latent distribution was CHOCO distributed (as there is no
reason to assume it would be different from Sample 1), treated as
continuous CHOCO-distributed data, despite the fact that it was
collected on a 7-point scale. The motivation was to see if the model
could extrapolate and reliably estimate all the parameters based on a
limited number of data points but means in our case that, effectively,
each Beta distribution was extrapolated from the frequency of only three
unique values. While this is far from ideal, the CHOCO model still
proved insightful, although it might have impacted its ability detect
finer-grain changes, including quadratic relationships.

@@

@@

\bibitem[\citeproctext]{ref-eagly1991beautiful}
Eagly, A. H., Ashmore, R. D., Makhijani, M. G., \& Longo, L. C. (1991).
What is beautiful is good, but\ldots: A meta-analytic review of research
on the physical attractiveness stereotype. \emph{Psychological
Bulletin}, \emph{110}(1), 109.

\bibitem[\citeproctext]{ref-fiske2002model}
Fiske, S. T., Cuddy, A. J., Glick, P., \& Xu, J. (2002). A model of
(often mixed) stereotype content: Competence and warmth respectively
follow from perceived status and competition. \emph{Journal of
Personality and Social Psychology}, \emph{82}(6), 878.

@@

\bibitem[\citeproctext]{ref-nakamura2014attractive}
Nakamura, K., \& Kawabata, H. (2014). Attractive faces temporally
modulate visual attention. \emph{Frontiers in Psychology}, \emph{5},
620.
